---
title: Distinct functional responses of consumers and their producers to climate drive mutualistic network asymmetry
subtitle: Functional Asymmetry in Ecological Networks
author:
  - name: Gabriel Munoz
    #orcid: 0000-0002-0760-5497
    corresponding: true
    email: gabrielmunoz1891@gmail.com
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Concordia University
  - name: Paul Savary
    #orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - Concordia University
  - name: W. Daniel Kissling
    #orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - University of Amsterdam
  - name: JP Lessard
    #orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - Concordia University
keywords: 
  - Climate change
  - Frugivory
  - Interaction diversity
  - Network specialization
  - Seed dispersal
abstract: |
  **Aim:** Functional traits are often used to infer the ecological processes that determine the composition of species assemblages. However, most trait-based approaches to infer community assembly processes focus on a single trophic level. Owing to the matching of traits facilitating interactions between resource and consumer assemblages, the functional trait diversity of different trophic levels is expected to covary in space. However, differential response of consumers and producers to environmental gradients can cause a decoupling of functional diversity between trophic levels, which we coin functional trophic asymmetry. Here, we develop a metric to quantify functional trophic asymmetry (FTA) and use it to infer the processes underpinning multitrophic community assembly, and explore the role of these processes in shaping the topology of ecological networks. 
  
  **Location:** Neotropics.
  
  **Time period:** Present.
  
  **Major taxa:** Mammalian frugivores and palms.
  
  **Methods:** We use digitally available data on the functional traits, pairwise mutualistic interactions, and geographic distributions of consumers (mammalian frugivores) and their producers (palms) to quantify functional trophic asymmetry for species occurring in the Neotropics. To cover major data gaps between species-level trait and interaction data, we train machine learning models to generate synthetic network data. We then use linear regression models to relate functional asymmetry to variation in climate, and to assess the influence of functional asymmetry on network specialization.
  
  **Results:** Our approach generated probabilistic networks across 1,072 grid cells in the Neotropics, revealing networks with a clearly defined modular structure and substantial differences in their functional richness across trophic levels. Functional trophic asymmetry increases from regions of the Neotropics with low precipitation seasonality to regions with high precipitation seasonality. Along this same climatic gradient, network specialization is positively related to functional trophic asymmetry.
  
  **Main conclusions:** Our results further suggest that mutualistic interactions between palms and mammals are mediated by matching traits and taxonomic overlap as a key assembly process at the regional scale. We conclude that increased warming and seasonal shifts in precipitation caused by global climate change could disproportionately impact specialist species while increasing functional trophic asymmetry in ecological networks.
  
plain-language-summary: |
  This study explores how differences in functional traits between co-occurring species assemblages of consumer and producers affect ecological network assembly in the Neotropics. By focusing on mammalian frugivores and palms as a case study, we developed a novel metric functional trophic asymmetry (FTA) to measure mismatches in trait diversity between trophic levels. Using machine learning and statistical models to agument spatially sparse data on species interactions, we found that FTA increases in areas with high precipitation seasonality and leads to more specialized ecological networks. The results suggest that climate change could disproportionately impact specialist species and increase FTA, altering the structure of species interactions in these ecosystems.
date: today
bibliography: references.bib
citation:
  container-title: Manuscript in Progress
number-sections: true
---

```{r}
#| echo: false
#| include: false
#| warning: false 
install.packages('gtsummary')
install.packages('broom')
install.packages('broom.helpers')
install.packages('cardx')
install.packages('Matrix')
install.packages('car')
install.packages('parameters')
library(gtsummary)
```


# Introduction

Ecologists often examine patterns of functional trait diversity to
investigate community assembly processes [@ackerly2003community;
@kraft2015community]. To date, however, trait-based approaches in
ecology often focus on a single trophic level, whereas approaches that
consider multiple trophic levels remain rare [@lavorel2013novel;
@seibold2018necessity]. An approach that considers processes operating
within and between trophic levels is necessary to better understand the
assembly of multitrophic communities[@allesina2008general;
@marjakangas2022trait; @saravia2022ecological]. Moreover, considering
trophic interactions while studying community assembly could shed new
light on processes underpinning ecological networks
[@allesina2008general] .

Classical approaches to study community assembly rely on the concept of
environmental filtering, sorting or selection, where density independent
conditions constrain the functional richness of species assemblages
[@laliberte2010distance; @villeger2008new;
@hillerislambers2012rethinking; @kraft2015community]. Functional
richness refers to the variability and relative frequency of different
functional traits observed in a community. It is often used to estimate
the strength of selection imposed by the environment
[@kraft2008functional; @kraft2010functional; @kraft2015community]. High
functional richness can indicate weak environmental selection whereas
low functional richness can indicate strong selection
[@kraft2008functional; @halpern2008functional; @paine2011functional]. In
a multitrophic context, the effects of environmental selection can
cascade across trophic levels such that selection on consumer traits can
shape the functional richness of their resources, modulated by their
degree of reciprocal dependency or co-evolution [@lavorel2013novel;
@guzman2019towards]. Moreover, the same environmental gradient could
exert selective pressures of different strength on communities at
distinct trophic levels [@marjakangas2022trait]. Differences in the
strength of selective pressure among trophic levels could then possibly
constrain the structure or topologies of trophic networks
[@bluthgen2007specialization; @schleuning2012specialization;
@dehling2020similar].

Inferring the relative strength of environmental selection between
trophic levels requires using high-dimensional approaches that are able
to deal with sparse observations for a large number of species
[@rohr2010modeling; @strydom2022food]. We introduce the concept of
functional trophic asymmetry (FTA), which allows inferring the relative
influence of environmental selection and trait matching on the
composition of multitrophic assemblages [(Figure 1)](Figure%201). FTA is
the difference in the richness of interaction-relevant traits between
trophic levels in a multitrophic network. FTA can occur because traits
mediating species interactions (i.e., interaction niches) across trophic
levels can also mediate the responses of species to their abiotic
environment (i.e. environmental niches) [@mccain2014body;
@moretti2009combining; @nagy2018arthropod; @dehling2020similar]. As an
example, plant seed size determine the outcome of animal-mediated seed
dispersal [@donoso2017complementary; @donoso2020downsizing] as well as
physiological limits, such as tolerances of plant seedlings to
desiccation [@hoekstra2001mechanisms]. High FTA could indicate
differences in the strength of environmental selection over the
interaction niches of distinct trophic levels within a multitrophic
species assemblage. Alternatively, low FTA could indicate that the
strength of the environment selection shaping interaction niches is
similar between trophic levels, e.g. equally weak or equally strong
[@marjakangas2022trait]. When interactions between producers and
consumers are mutualistic, low FTA could also emerge under strong trait
matching and therefore indicate the influence of trait-coevolution
during multitrophic community assembly [@dehling2014functional;
@albrecht2018plant]. By studying spatial variation in functional trophic
asymmetry along environmental gradients, we could possibly identify the
conditions promoting environmentally versus cross-trophic interaction-
driven community assembly [@schleuning2020trait; @bello2023analyzing].

Frameworks linking multitrophic functional diversity to network topology
along broad-scale environmental gradients are crucial to understand the
effects of global change on biodiversity and ecosystem function
[@bello2023analyzing; @dehling2018structure; @schleuning2020trait;
@schleuning2014ecological; @schleuning2023animal]. Functional responses
of consumer and producer assemblages to climate influence functional
richness at the level of the multitrophic community
[@garcia2018frugivore]. Because some of these traits are involved in
interactions across trophic levels, the filtering of traits along
environmental gradients could constrain the number of unique
interactions and therefore, network topology [@marjakangas2022trait;
@emer2020seed; @albrecht2018plant]. As an example, low multitrophic
functional richness could influence emergent patterns in network
structure such as the specialization of multispecies interactions by
limiting the relative availability of interaction partners across
trophic levels [@bluthgen2007specialization; @bluthgen2006measuring].
While high levels of network specialization represent networks
predominantly made of “one-to-one” interactions, low levels of network
specialization represent networks with species showing predominantly
“one-to-many” interactions [@bluthgen2006measuring;
@bluthgen2011functional] (Figure 1B). One highly expected outcome is
that when functional trophic asymmetry is high, networks will have low
specialization. For example, take a plant community exhibiting a low
richness of flower displays and which is associated with a bee community
(pollinators) exhibiting a wide variety of proboscis lengths. These
plants are unlikely to form “one-to-one” interactions with only a subset
of bee species that have matching proboscis length. Otherwise,
non-matching pollinators would have no food resource and be extirpated.
By partitioning deviations from expected FTA and network specialization
relationships with null models, one can separate the relative influences
of processes operating between trophic levels (e.g. trait matching) and
those within trophic levels (e.g. environmental selection) in network
assembly [@marjakangas2022trait]. However, the relationship between
network specialization and functional trophic asymmetry has not been
fully explored.

Preserving mutualistic interactions between palms and their mammalian
frugivores is important to sustain biodiversity and ecosystem function
in the tropics [@bogoni2020extent; @marques2022mutualism]. Mammalian
frugivores facilitate the dispersal of palm fruits, which helps to
prevent local extinctions amid disturbance and to maintain biodiversity
in these ecological networks [@acevedo2020sampling;
@messeder2020searching; @dehling2018structure]. Exploring patterns of
co-variation in the functional richness of palm and frugivore
assemblages along environmental gradients is a first step in this
direction. Here, we ask (1) which climatic variable(s) best explains
variation in the functional richness of palms vs mammal frugivores, (2)
whether differences in these relationships lead to functional trophic
asymmetry, (3) where does functional asymmetry peaks in the Neotropics
and (4) whether asymmetry relates to network specialization.


![Figure 1: This conceptual model illustrates the dynamic relationship between
functional diversity metrics—specifically Functional Richness (FR) and
Functional Trait Asymmetry (FTA)—and environmental gradients within
ecological networks. The left panel of **Figure A** visualizes the
variation in FR for producers (depicted in green) and consumers
(depicted in orange) along an environmental gradient. As the
environmental gradient intensifies (e.g., through changes in
temperature, precipitation, or habitat fragmentation), FR for both
producers and consumers generally declines. However, this decline can
occur at different rates, leading to two scenarios: (1) **Differential
Decline in FR**: If consumer FR declines more sharply than producer FR,
a substantial increase in Functional Trait Asymmetry (FTA) occurs. (2)
**Parallel Decline in FR**: Alternatively, if both producer and consumer
FRs decline at a similar rate, FTA remains relatively constant along the
gradient. This scenario indicates a balanced impact of environmental
changes across trophic levels, preserving the relative functional
relationship between producers and consumers. **Figure B** shifts focus
to the implications of changing FTA on network specialization—a measure
of how distinct or generalized interactions are between producers and
consumers within ecological networks. **Higher Specialization**: This
scenario, shown on the left side of the gradient, involves more distinct
producer-consumer interactions, where specific consumer species interact
with particular producer species. This high specialization often
correlates with low FTA, where the functional traits between interacting
species are closely aligned. **Lower Specialization**: As environmental
stress increases, leading to a higher FTA, the network may shift towards
lower specialization. In this state, interactions become more
generalized, with consumers utilizing a broader range of producers due
to the loss of specific functional
traits.](Main_figures/00_Figure1.png){fig-align="center" #fig:fig1}


# Methods

In this study, we investigated the variation of functional trophic
asymmetry and network specialization along climatic gradients by
gathering species-level information on traits, interactions, and
distributions from literature, museum specimens, and field collections.
The collected data were then processed to create synthetic networks
across gridded regions of the Neotropics and to calculate functional
trophic asymmetry and network specialization variables, followed by
using climate data to explain the variation in these variables across
the Neotropics (Figure S1).

## Study system

We focused on multitrophic communities of Neotropical palms and their
mutualistic, seed dispersing, mammalian frugivores. Palms
(Plantae:Arecaceae), being a keystone plant family in tropical regions
[@kissling2012towards; @onstein2017frugivory], provide fruit resources
to a wide variety of vertebrate frugivores, including birds and mammals
[@munoz2019synthesis; @zona1989review]. Frugivore mammals
(Animalia:Mammalia) are among the most important palm-seed dispersers,
particularly over long distances. Most frugivore mammals feeding on
palms are seed eaters and pulp eaters, dispersing palm seeds mostly via
ectozoochorus dispersal [@messeder2020searching]. Importantly,
frugivory-related traits have notably underlain palm diversification and
played a key role in the evolution of palm traits
[@onstein2014diversification; @onstein2017frugivory]

## Data collection

### Species level geographic distribution data

We obtained binary species distribution data (present/absent) on palms
from the geographic range maps of [@bjorholm2005environmental] and on
mammals from the IUCN (International Union for the Conservation of
Nature) data portal. To generate local gridded multitrophic species
assemblages across the Neotropics, we intersected the species-level
range maps with a spatial grid where each grid cell represented every 1
by 1 degree latitude and longitude change along the extent of the entire
Neotropics. We then listed all palm and mammal frugivore species
co-occurring in each grid-cell as our grid-cell level multitrophic
assemblage.

### Species level trait data

We collected species-level multitrophic trait data related to the
physiological tolerance of palms and frugivorous mammals to the abiotic
environment and to their mutualistic interactions. For palms, we
extracted data from the PalmTraits 1.0 dataset
[@kissling2019palmtraits]. We collected data on growth form, maximum
stem height, and average fruit length. For frugivorous mammals, we
obtained trait data from the EltonTraits 1.0 database
[@wilman2014eltontraits]. We selected data on body mass, diet, and daily
activities. Diet data from the EltonTraits 1.0 database is coded as
percentage use distribution across ten diet categories. We excluded from
our analysis species without fruit in their diet. Activity was coded as
a dummy variable with three categories (Diurnal, Crepuscular,
Nocturnal). Finally, body mass was coded as a numerical variable in kg.
We excluded bats from the analysis as almost no Neotropical bat species
is feeding on palm fruits [@messeder2020searching]. In total, we subset
from this dataset the species with available distribution range map,
that is 494 palm species and 737 mammal frugivores.

### Pairwise species level interaction data

We used data on seed dispersal interactions between palms and mammals
for the Neotropics, originating from recollections of seed dispersal
records found in the published literature and interaction records are
recorded at the species level [@munoz2019synthesis]. Each pairwise
species interaction record reflects where an article mentions the fruit
or the seed of a palm being dispersed, carried or defecated by a
frugivorous mammal. Interaction records collected in this database were
previously vetted to reflect effective seed dispersal interactions,
while avoiding those that reflect mere seed consumption (vetting
criteria found in: [@munoz2019synthesis]). In total, we gathered a total
of 581 interaction records between 69 palms and 111 frugivore mammals.

### Grid-cell level environmental data

We used bioclimatic variables from WorldClim [@fick2017worldclim] to
represent large-scale spatial and temporal variation of climate in the
Neotropics. Specifically, we used mean annual temperature (BIO01), total
annual precipitation (BIO12), temperature seasonality (BIO04) and
precipitation seasonality (BIO15). Using a moving window, we compute
simple averages for every set of bioclimatic records at each grid cell,
thereby re-scaling the spatial resolution of bioclimatic variables to 1
by 1 degree grid resolution from their original resolution (1 by 1 km2)
to match the spatial resolution of our grid cell species-level data.

### Continental level biogeographical data

The Neotropics is a region with a rich evolutionary history which
significantly influenced patterns of species colonization and extinction
across neotropical plant and animal species [@whiteman2015into;
@antonelli2011there] . The biogeographic regionalization patterns of
[@morrone2014biogeographical] distinguish seven major biogeographic
regions (i.e., biogeographic dominions), we use them to delineate the
spatial extent of species pools when simulating network assembly
processes.

## Statistical analyses

### A probabilistic continental metaweb

Here, we fitted latent variable models that vary in their assumptions to
estimate interaction probabilities from observed binary data on species
interactions. Specifically: the stochastic block model (SBM), the
connectance model, the trait-matching model, and the matching-centrality
model [@terry2020finding]. The SBM assumes that ecological networks are
modular, with species interacting more within their groups, and outputs
an two incidence matrix for palm and mammal species group affiliations,
and a squared matrix (Theta matrix) for interaction probabilities. The
connectance model posits that interactions of specialist species are
subsets of those of generalist species, optimizing connectivity scores
to recreate observed network patterns. The trait-matching model assumes
non-random species interactions determined by trait differences,
optimizing parameters along latent-trait axes. The matching-centrality
model combines connectivity scores and latent-trait axes
[@terry2020finding]. We fitted these models to our available interaction
data and selected the model that best predicted the observed continental
pattern of seed dispersal interactions. Using Youden’s J as a metric
that balanced model sensitivity and specificity [@poisot2023guidelines],
we find that SBM was the best supported model (Figure S2 A-C).
Additional details about the model assumptions are explained in
***Supplementary Text S1**.*

### Downscaling the continental metaweb to generate grid-cell level networks

The digitally availability of primary biodiversity data on palms and
their mammalian frugivores was imbalanced, with well covered data in
terms of distribution ranges, followed by well covered data on species
traits, to a limited number of interaction records. Therefore, to
downscale our initial metaweb to include interactions between between
every potentially co-occurring palms and mammal frugivore at every
gridcell of the Neotropics, we use a two fold approach. First, we
employed multinomial logistic regression models that aimed to predict
the species level SBM model results (i.e. group affiliations) from
species-level trait data. We justify the choice of multinomial logistic
regression models as these can handle the prediction of non-binary
outcomes, that is in our case, the labeling of SBM groupings per
species. We fitted separate multinomial models for palms and mammal
frugivores using a label backpropagation algorithm and a neural network
engine, with 75% of the data allocated for training and the 25%
remaining for testing. We used neural networks because they are useful
when dealing with multicollinearity, as they can learn complex and
non-linear relationships and interactions among predictor variables.
This allowed us to separate the relative importance of distinct matching
traits on SBM group affiliations. We extracted variable importance
scores based on the combinations of the absolute values of the best fit
model weights [@gevrey2003review]. Second, we considered as local
pairwise species interaction probabilities as the product of the values
from the Theta matrix from the SBM model that represent the latent
interaction probabilities between species pairs within and between
groups multiplied by their probability of co-occurrence (POC) in a
gridcell. To represent species’ co-occurrence probabilities, we used the
reciprocal distance between the centroids of species pair ranges within
the grid-cell, divided by the sum of its range areas within the
grid-cell. This implied that within each grid cell, species with closer
range centroids and larger cumulative areas are more likely to co-occur
and interact. This approach allowed us to recreate synthetic
probabilistic plant-mammal frugivore networks for each grid-cell across
the Neotropics, while accounting for the heterogeneity of species ranges
within each grid.

### Measuring Functional Trophic Asymmetry (FTA) and Network Specialization (H2')

We computed Functional Trophic Asymmetry (FTA) from the results of the
SBM model fit. Specifically, from the matrices representing the
incidence of palm or mammal frugivore species in one of the SBM
groupings. Thus, as our measure of functional richness, we calculated
the number of species of each taxon across SBM groups per grid. Because
we had differences in the total number of palm and mammal species across
grids, we normalized species counts within gridcells. We then computed
the absolute difference between trophic levels to obtain a measure of
FTA for each combination of SBM groups. Because we had the potential for
each palms and mammals species at each grid to become affiliated to any
of 7 SBM groups, and to interact with any species of the opposite
trophic level within and between groups, we obtained a total of 49
independent FTA measures for each gridcell, one for each palm-mammal
group combination.

We measured network specialization at each grid cell using the metric
H2’. H2’ is a network-level index that describes the degree of
specialization of interactions between species
[@bluthgen2007specialization]. High values indicate networks that are
more specialized, meaning that specialist species from one trophic level
interact with specialist species from the opposite trophic level. Low
H2’ values indicate networks among generalists, meaning that there is a
low specificity of interactions between species across trophic levels.
Because inferred networks varied in their network size (i.e., number of
unique interactions between palms and mammals), we rarefied the
computation of H2’to networks of the same absolute size per gridcell,
resampling networks to the same number of pairwise interactions (100) at
each grid-cell 999 times [@terry2020finding]. We selected the median of
this H2’ distribution as our gridcell level measure of network
specialization.

We assessed the deviance of ‘observed’ FTA and H2’ from null models that
simulated stochastic community assembly processes [@dormann2009indices].
We created these null models by constructing networks of interactions
between a randomized set of palms and mammals for each gridcell and
computing expected values of FTA and H2’. To do this for a given
gridcell, we identified their biogeographic domain (from
[@morrone2014biogeographical]) and randomly sampled 10 palm and mammal
species, ensuring each species is selected only once. To this subsampled
species set, we used the same procedures to calculate FTA and H2’ and
replicated the process 999 independent times to obtain a distribution of
expected FTA and H2’ values. Finally, we computed the deviance between
observed and expected values with Z-scores.

### Assessing the influence of climate in Functional Trophic Asymmetry

We fitted linear regression models to investigate the influence of
climate to Functional Trophic Asymmetry. We used FTA z-scores as
response variables and the four climatic predictor variables: Mean
Annual Temperature (Temp), Total Annual Precipitation (Prec),
Temperature Seasonality (TS), and Precipitation Seasonality (PS) as
predictor variables. In addition, we included the identity of the SBM
palm-mammal group combination as an interaction term. Prior to fitting
the models, we scale-transformed the climatic variables such that they
all share the same mean but only differ in their standard deviation.

### Assessing the influence of Functional Trophic Asymmetry in Network Specialization

We fitted linear regression models to investigate the effects of
Functional Trophic Asymmetry in Network specialization. We used the
average (mean) and standard deviation (sd) of FTA across all gridcells
and SBM combinations as our predictor variables. By including both as
predictors, the model accounts for not only the overall asymmetry but
also its distribution among SBM groupings. We used the rarefied gridcell
records of H2’ as our response variable. To asses model significance, we
compared the estimates of this model to a distribution of expected model
coefficients from fitting the same linear model for each of the null
network assembly replicates.

# Results

## The structure of mutualistic networks between palms and their mammalian frugivores across the Neotropics

The SBM model organized the continental set of palm and mammal species
into a matrix sorting species into seven SBM groups and an accompanying
matrix that quantified interaction probabilities between and within
groups. In our model, there was a 1.74 fold increase in the likelihood
of interactions between species within SBM groups rather than between
SBM groups (Figure 2A). For palms, important trait variables that
predicted SBM group affiliation were palm fruit length, maximum stem
height, and palm growth form (*Acaulescent* or *Erect*) (Figure S3-A).
For mammals, important trait variables to predict species group
affiliations were mammal activity, the logarithm of the *body mass*, and
percentage of frugivory on diet as predictor variables (Figure S3-B).
These analysis revealed that SBM group assortments represent communities
of palms (Figure S4) and mammal frugivores (Figure S5) with unique trait
combinations within the high-dimensional multitrophic spectrum of
interaction-relevant traits. Our approach to downscale a continental
metaweb was able to generate probabilistic networks between palms and
mammalian frugivores for 1,072 grid cells acrosss the Neotropics. Each
network had in average nine interaction partners per species (sd =
5.62), 27 mammals per grid cell (sd = 14.97), and 11 palms per grid cell
(sd = 11.42).




![Figure 2: Spatial distribution of **(A)** FTA z-score (mean) and **(B)** H2'
network specialization index across the Neotropics. Panel A depicts the
geographic variation in the mean z-scores of the Functional Trophic
Asymmetry (FTA). Areas with higher z-scores indicate greater asymmetry
in the richness of interaction relevant traits between trophic levels,
whereas lower scores suggest more balanced trait distributions. Panel B
presents the spatial distribution the degree of species specialization
of interactions within ecological networks. This panel reveals
significant latitudinal gradients, particularly marked in the Andes and
the Southern
Cone.](Main_figures/00_network_maps.tif.png){fig-align="center"  #fig:fig2}



## Differences in trait-climate relationships across trophic levels lead to Functional Trophic Asymmetry

FTA z-scores [(Figure 2B)](#fig2) indicate that in average FTA was less than the
FTA expected from a null assembly model.  Out of the 49 SBM group combinations,
those with less than expected asymmetries are generally between palms in
groups 1, 2, and 5 and mammals in groups 6 and 3. Conversely,
combinations with significantly greater than expected asymmetries are
generally found between palms in SBM groups 6 and 4 and mammals in SBM
groups 1, 2, and 5. FTA peaks in tropical and mountainous regions while
being lower in more temperate areas of the Neotropics, and remains
stable throughout tropical and humid regions (Figure 2A). The results of
our linear model shows significant differences in FTA among SBM groups.
In addition, it shows a relationship between the spatial variation of
climate and FTA. However, the significance and direction of effects vary
as a function of the SBM group combinations (Figure 3B, Table 1). While
the overall fixed model estimates do not show significant effects of
climatic variation in FTA across all SBM group combinations \[Temp (Mean
annual temperature) ( 𝞫 =0.0051, p=0.859), Prec (Total annual
precipitation) ( 𝞫 =−0.0364 p=0.384), TS (Temperature Seasonality) (𝞫
=−0.0180 p=0.746), and PS (Precipitation Seasonality) (𝞫 =−0.0507
p=0.250)\]; for specific group combinations, FTA is significantly
predicted by climate, particularly by changes in Precipitation
Seasonality (Figure S6). Approximately 59% of the variability in the FTA
z-score is explained by the linear model with our climate predictors as
fixed effects and SBM group identity as interaction terms (adjusted
R-squared = 0.59).

On one hand, group combinations exhibiting a positive relationship
between FTA and precipitation seasonality are characterized by a
relatively stable functional richness of mammals across gradients of
precipitation seasonality, while the functional richness of palms
increases under these conditions (Figures S6, S7). A significant portion
of this pattern can be attributed to turnover within the palm community
of SBM group 6, which predominantly consists of small to medium-sized
drupes or berries. These palms, which belong to genera such as
*Acrocomia*, *Allagoptera*, and *Astrocaryum* are typically of tropical
origin and display diverse growth forms, including both erect and
acaulescent. Conversely, group combinations showing a negative
association between FTA and precipitation seasonality are primarily
driven by patterns in the palm community of SBM group 5 (Figures S6,
S7). In this scenario, while the functional richness of mammals remains
stable, the functional richness of these palms, generally characterized
by species with erect growth forms and large fruits, increases in
regions with low precipitation seasonality.

![Figure 3: **(A)** Heatmap depicting the interaction probability among various
multitrophic communities across different SBM (Stochastic block model)
groups. The intensity of red shading correlates with the strength of
these interactions, where darker shades signify higher probabilities of
interaction between species within or between SBM groups. **(B)** The
left panels present geographical maps illustrating the distribution of
FR for both mammals and palms, identifying regions with distinct levels
of functional diversity for a given multitrophic combination of species
at one SBM group pair. The right panels illustrate the significant
relationship observed between FTA and precipitation
seasonality.](Main_figures/00_network_fr.png){fig-align="center"  #fig:fig3}


## Functional Trophic Asymmetry increases Network Specialization

We found overall, palm-mammalian frugivore networks had low
specialization, H2' ranged from 0.12 to 0.25. In addition, there was a
significant relationship between aggregated gridcell records of
Functional Trophic Asymmetry and H2' metric. Specifically, FTA z-scores
increased H2’ by 3.28 units per mean unit increase across all SBM
groups. Additionally, the standard deviation of FTA z-scores decreased
H2’ by 1.01 units per unit increase across all SBM groups (SD FTA).
While this linear model explained only around 8.7% of the total spatial
variability in H2’ (adjusted R-squared = 0.087), the results of our
simulations show that the intercept and slope estimates are
statistically significant from null assembly models (Figure S7).




# Discussion

By making use of digitally available primary biodiversity data
at broad scales, our approach provides novel insights into the assembly of mutualistic networks. By identifying species clusters of palms and mammals with
similar interaction patterns, we  offer a more nuanced understanding
of network assembly dynamics within these multitrophic communities. The distribution of Functional Trophic Asymmetry (FTA) across different climatic
regions of the Neotropics underscores the importance of contemporary
climate change in shaping interaction niches across species'
distributional ranges, with precipitation seasonality emerging as a
key predictor of FTA for certain SBM palm and mammal group combinations,
where higher FTA increase overall network specialization.

The delineation of interactions into distinct groups indicates that, at
a continental scale, the network of seed dispersal interactions is
highly compartmentalized. This finding aligns with previous work
reporting significant modularity in palm seed dispersal interactions
across the Neotropical continent [@munoz2019synthesis]. In our study, we
extend this understanding by identifying palm growth form and mammal
foraging activity periods as key palm and animal traits driving this
broad scale modularity through changes in functional richness. Clusters of similarly interacting species obtained from SBM models such as ours
have been shown to reflect spatio-temporal segregation of species assemblages and the
differentiation of interaction niches [@durand2023trait]. The observed distribution
of interaction probabilities within and between SBM groups highlights
multitrophic assemblages being central to the entire metaweb, notably
the palm and mammal species in SBM groups 5 and 6, which are those with
extensive distributional ranges and more generalist diets. Conversely,
it also identifies peripheral groups, such as species in SBM groups 1-3,
which exhibit narrower interaction niches. These peripheral species
assemblages may correspond to those species with restricted ranges
or specialized diets. Indeed, SBM groups 5-6 include larger, nocturnal
mammals that feed on palms with varying fruit sizes and growth forms,
ranging from erect to acaulescent, whereas SBM groups 1-3 predominantly
consist of smaller, diurnal mammals that consume small fruits from erect
palms.

Our linear model highlights Precipitation Seasonality as a key factor predicting the Functional Trophic Asymmetry (FTA) in key SBM groupings. This suggests
that spatial variation in interaction niches emerges from the
differences in the strength of environmental selection between trophic
levels. Such uneven pressures result in multitrophic assemblages where
not all co-occurring trait combinations align optimally with the global preferences
between species interaction partners, thus contributing to promote
structural diversity in seed dispersal networks across Neotropical
sites. For example in SBM groups 1-3, mammalian frugivores, which can exhibit a broad
tolerance for varying levels of precipitation seasonality, may
predominantly consume small-fruited palms in regions with high
precipitation seasonality. This is likely because palms with smaller
fruits are better adapted to thrive under conditions of variable water
availability. Consequently, increased aridity (e.g. due to climate change) is expected to result in increased FTA, driven by the restricted functional richness of
available resources for consumers. In SBM groups 5-6, palms with larger fruit
sizes and predominantly erect growth forms consumed by large sized
frugivores might reflect adaptations to attract high quality animal seed
disperses in environments where consistent water availability promotes
higher habitat complexity and plant competition. Consequently, increased aridity is expected to result in increased FTA, but this time driven by the restricted functional richness of available consumers relative to producers. In both cases, FTA imposes limits over trait matching between partners of distinct trophic levels and the partitioning of interaction niches within species of the same trophic level.

The a low degree of specialization observed for palm-mammal frugivore seed
dispersal networks is not entirely unexpected given the historical
ecological shifts that these species have experienced in the Neotropics.
Unlike other major tropical regions, such as the Afrotropics and
South-East Asia, the Neotropics had significant post-Pleistocene
extinctions, which heavily altered mammalian communities. These
extinctions, which impacted disproportionately to mammal megafauna,
likely exacerbate competition and extinction rates among their
large-fruited palm partners, particularly in those regions with high
mammal diversity and endemism. [@webb2006great; @whiteman2015] into
Concurrently, the radiation of small-fruited palm taxa
[@onstein2014diversification; @onstein2017frugivory] further contributed
to the decoupling of trait diversity between palms and mammals and the
re-structuring of seed dispersal networks across trophic levels in the
Neotropics. Without this trait decoupling due the co-evolutionary
dynamics between palms and mammal frugivores may have remained
unaltered, fostering otherwise more specialized networks.

The positive estimate for mean FTA suggests a strong positive influence
of FTA on H2’, thus promoting network specialization. In contrast, the
negative estimate for FTA standard deviation (SD) indicates that a large
variation in FTA within SBM group combinations significantly reduces
H2', implying more generalist interactions. This suggest, that while
overall higher FTA promotes network specialization, an inconsistent FTA
across SBM groups can disrupt this pattern, making network structure
less predictable from multitrophic trait patterns. Previous research in
insular regions where trophic communities are often subject to
asymmetric filtering strengths have shown that the specialization of
mutualistic networks can indeed be resilient to shifts in the degree of
ecological filters between trophic levels. Particularly, if the networks
are well supported by generalist species [@schleuning2012specialization
; @schleuning2016ecological]. It is therefore critical to consider
changes in the underlying network structure when evaluating potential
effects that climate-driven species or species trait losses have in
ecosystem function [@acevedo2020structure; @schleuning2020trait;
@schleuning2023animal, @classen2020specialization]

We validated the statistic significance of the model parameters through
null simulations of network assembly,. However, it is noteworthy that
our null models are limited to presence/absence permutations. Therefore,
the potential influence of varying species relative abundances across
sites was not considered in this study. It is well established that
population-density dependent effects can modify interaction
probabilities from trait matching and range co-occurrences [@pena2023abundance; @donoso2017complementary; @mcfadden2022global]. 
Future studies should aim to incorporate species abundance estimates
when defining interaction probabilities. However, despite ongoing
efforts, such data is not readily available for multiple species at
different trophic levels, and at broad spatial scales. Our results help
elucidate mechanisms that contribute towards a broader theory of
multitrophic community assembly. However, given the current limitation
in obtaining reliable species abundance data, we advise caution when
utilizing our data pipelines for making finer-scale spatial predictions
of network assembly and interaction diversity. Similarly, the use
biogeographic domains to delineate the species pools in the construction
of our null models respects ecological theory and considers both
historical and regional species pools [@mcfadden2022global;
@carstensen2013introducing; @cornell2014species]. However, one critical
assumption of this approach is that species within each biogeographic
dominion are all equally likely to colonize a given community
[@lessard2012inferring] While this assumption may simplify the
computational complexity of our algorithms, future work could penalize
dispersal distances with landscape level variables such as the slope
terrain [@preau2022dispersal; @schlagel2020movement]. Finally, the
clustering of interactions captured by the SBM may be also capturing
reveal relatively unexplored drivers of interaction assembly, such as
phylogenetic niche conservatism [@ackerly2003community;
@wiens2005niche; @pyron2015phylogenetic], where closely related species
retain similar ecological traits over evolutionary time, and/or the
widespread extinctions of large mammals frugivores and the rapid trait
speciation of small fruited palms [@donoso2020downsizing;
@lim2020frugivore; @onstein2017frugivory] . Partitioning these aspects
and their relative influences in delineating SBM group assignments can
be future research avenues to better understand the underlying
mechanisms behind mutualistic interaction patterns in space and time.

Unraveling the complexity of network assembly processes across broad
spatial scales requires continous development and refining of
statistical workflows and biodiversity data pipelines that integrate
digitally available data from diverse sources [@kissling2012towards;
@thuiller2024navigating]. Continued field collection, their open-release,
and data standardization is crucial if we want to expand our capacity to
training models capable of predicting ecological network structures at a
global scale with a high degree of spatial and/or temporal
resolution[@poisot2016describe; @poisot2021global]. While many of these
modelling approaches are in early development, their success hinges on
the availability and reliability of foundational data. We have
demonstrated that with sparse data across large scales, novel models can
indeed be developed. Filling global data gaps and leveling data
completeness across species ranges, traits, and interactions can
certainly enhance predictive accuracy. While our case study has a
history of digitally available natural history data on both palms and
mammal frugivores, such comprehensive datasets are not available for
many other tropical clades. We thus advocate for increased awareness,
action, and funding for global biodiversity programs that collect and
digitize natural records, thus enabling the continous development of
large-scale spatially explicit predictive models that inform about
mechanisms of network assembly, and their relationship to climate and
ecosystem function.

# Tables

Table 1: Linear regression analysis of the relationship between
Bioclimatic Predictors (Mean Annual Temperature, Total Annual
Precipitation, Precipitation Seasonality, Temperature Seasonality) and
Functional Trophic Asymmetry within SBM groups (p1m1 - p7m7).
Statistically significant effects (p\<0.05) are marked with (\*). Beta
correspond to the variable effect coefficient in the model and SE
correspond to the standard error.

```{r, echo = FALSE, warning=FALSE, message=TRUE}


mymod <- readRDS("00_Data/04_models/lm_asym_all.RDS")

gtsummary::tbl_regression(mymod, 
                          label = list('lab' = 'SBM group', 
                                       'scale(Temp)' = 'Mean Annual Temperature', 
                                       'scale(Prec)' = 'Total Annual Precipitation','scale(TS)' = 'Temperature Seasonality', 'scale(PS)' = 'Precipitation Seasonality')) |>
  bold_p(t = 0.05)|>
  bold_labels() |>
  gtsummary::italicize_levels() |> 
  gtsummary::add_significance_stars()

```

Table 2: Linear regression analysis of the relationship between
Functional Trophic Asymmetry and Network Specialization. Statistically
significant effects (p\<0.05) are marked with (\*). Beta correspond to
the variable effect coefficient in the model and SE correspond to the
standard error.

```{r, echo = FALSE , warning=FALSE}
mymod <- readRDS("00_Data/04_models/obs_model1.RDS")

gtsummary::tbl_regression(mymod, label = list('mean_fta' = "FTA (mean)",
                                              'sd_fta' = 'FTA (sd)')) |>
  bold_p(t = 0.05)|>
  bold_labels() |>
  gtsummary::italicize_levels() |>
  add_significance_stars() |>
  add_global_p() 


```



# References 

:::{#refs}

:::


# Supplementary Material

## Supplementary figures

![Figure S1: A comprehensive workflow designed to integrate and harmonize primary
biodiversity data from digitally accessible datasets. In our case study,
these datasets are systematically categorized into three key components:
palm-mammal interactions, species traits, and geographical distribution.
The workflow is structured to address data imbalances through network
parametrization, which ensures that underrepresented data is
appropriately synthesized to generate a more balanced dataset.
Additionally, trait-linkage analysis is employed to examine
relationships between latent traits, facilitating a deeper understanding
of species interactions. Further, we examine the assembly of
multitrophic networks, which are constructed based on their matching
traits and the spatial overlap of species distributions. Ultimately,
these networks can provide insights on the drivers of network assembly
across diverse environmental
gradients.](Sup_figures/00_figure_data_pipeline.png){fig-align="center"}


![Figure S2: Evaluating distinct network structural models to predict mutualistic
interactions between palms and mammals in the Neotropics. **(A) Receiver
Operating Characteristic (ROC) Curves**: This panel illustrates the
performance of different network structural models in predicting
interactions across species pairs. Each curve is color-coded to
represent a distinct model, providing insight into the trade-off between
true positive and false positive rates across varying thresholds. **(B)
Boxplots of Model Accuracy**: This panel compares the accuracy of the
models, displaying the distribution of accuracy scores across multiple
datasets or iterative simulations. The variability, indicated by the
spread of the boxplots, highlights the consistency and reliability of
each model's predictions. **(C) Precision-Recall (PR) Curves**: This
panel focuses on the trade-offs between precision and recall (evaluated
with Youden's J) at different probability thresholds for each model. The
curves help to identify the most effective models in minimizing false
positives while maximizing true positives, especially in imbalanced
datasets such as ours where interactions are sparse. **(D) Heatmaps of
Predicted Adjacency Matrices**: This panel visualizes the predicted
interactions in the form of adjacency matrices for the four distinct
models examined in this study: the Trait Matching Model, Connectance
Model, Matching-centrality Model, and Stochastic Block Model.
Hierarchical clustering applied to species based on their predicted
interactions helps in elucidating the structural differences between
models, revealing patterns of species clustering and potential
interaction
modules.](Sup_figures/00_figure_data_metaweb_fit.png){fig-align="center"}


![Figure S3: **Variable Importance for Interaction Traits in Palm and Mammal
frugivores**. The left panel presents the relative importance of palm
interaction traits, where *Acaulecscent* and *Erect* growth forms are
identified as the most influential traits to define interactions with
mammals. Secondary traits such as *MaxStemHeight_m* and
*AverageFruitLength_cm* contribute moderately to delineate interactions.
The right panel shows the importance of mammal interaction traits in
shaping their interactions with palms, with activity patterns—nocturnal
and diurnal—emerging as the most critical. Crepuscular activity also
shows notable relevance. Among diet-related traits, scavenging and fish
consumption, alongside body mass, are recognized as important, though
their influence is comparatively
lower.](Sup_figures/00_Variable_importance.png){fig-align="center"}



![Figure S4: **Mammal trait representations across SBM groups (interaction
blocks).** **Dietary Traits (Left Panel):** The heatmap illustrates the
percentage distribution of various mammalian diet types across the
interaction blocks (SBM groups). Darker shades indicate a higher
prevalence of specific diet types within particular blocks. This
distribution reveals the trophic specialization of mammals and suggests
how different diet types are clustered or dispersed across ecological
interactions. **Activity Patterns (Middle Panel):** The second heatmap
focuses on mammalian activity data, categorized as Diurnal, Crepuscular,
and Nocturnal. Darker shades signify higher percentages, allowing for a
comparison of how activity patterns are distributed across the same
interaction blocks. This panel helps in understanding temporal niche
partitioning and its relationship to ecological interactions. **Body
Mass Variation (Right Panel):** The boxplot illustrates the distribution
of mammalian body mass (log-transformed) across the interaction
blocks.](Sup_figures/Trait_associations.png){fig-align="center"}


![Figure S5: **Palm trait representations across SBM groups (interaction blocks).**
The left and middle panels feature boxplots that display the
log-transformed average fruit length and maximum stem height,
respectively, for each interaction block. The right panel shows the
relative proportions of plant growth forms within each interaction
block. The color gradient represents the percentage contribution of each
growth
form.](Sup_figures/00_palm_trait_decomposition.png){fig-align="center"}



![Figure S6: **Predicted z-scores of FTA (Functional Trait Assemblage) as a
function of four climate variables and across combinations of seven
different palm (p1 to p7) and seven mammal (m1 to m7) SBM groups**. The
four panels represent: Mean Annual Temperature (top left), Total Annual
Precipitation (top right), Precipitation Seasonality (bottom left), and
Temperature Seasonality (bottom right). Solid lines indicate
statistically significant relationships (p \< 0.05), while dashed lines
indicate non-significant relationships (p ≥ 0.05). The different
interaction models are color-coded as indicated in the legend and the
facet title.](Sup_figures/00_climate_effects.png){fig-align="center"}


![Figure S7: **Frequency distribution of model estimates for the intercept and
Functional Trait Asymmetry (FTA) metrics**. The three panels display the
distribution of model estimates for the intercept (left), FTA mean
(center), and FTA standard deviation (right). Red vertical lines denote
the actual estimates obtained from the model, serving as a reference
point against the resampled or bootstrapped estimate distributions. The
histograms illustrate the frequency count of estimates derived from a
null model that reshuffled species co-occurrences. FTA mean is the mean
FTA z-scores for all SBM palm-mammal group combinations in a gridcell.
FTA sd is the standard deviation of the distribution of FTA z-scores for
all SBM palm-mammal group combinations in a
gridcell](Sup_figures/00_simulations_estimates.png){fig-align="center"}

## Supplementary text



**Supplementary text S1: Additional details about the structural network models used to predict species interactions in this study**

To capture probabilistic patterns from interactions observed at the continental scale, we tested four assembly models: the stochastic block model (SBM), the connectance model, the matching-trait model, and the matching-centrality model. Each model was evaluated based on its ability to predict interaction probabilities between palm and mammal species pairs.


- **1. Stochastic Block Model (SBM):**

  The SBM assumes that ecological networks exhibit a modular pattern, where subsets of species interact more within their groups of highly connected species (Terry and Lewis, 2020).

The model generates an incidence matrix, \( Z \), reflecting species-level associations to a group, and a squared matrix, $( \Theta )$, reflecting the interaction probabilities within and between groups. The likelihood \( L \) of the SBM is given by:

$$
L = \prod_{i,j} \left( \Theta_{Z_i Z_j} \right)^{A_{ij}} \left( 1 - \Theta_{Z_i Z_j} \right)^{1 - A_{ij}}
$$


where

- $( A )$ is the adjacency matrix of observed interactions,
- $( Z_i) (Z_j)$ are the group memberships of species \( i \) and \( j \), and
- $( \Theta_{Z_i Z_j} )$ is the probability of interaction between groups \( Z_i \) and \( Z_j \).



- **2. Connectance Model:**


The connectance model posits that the interactions of specialist species are subsets of those of generalist species. It optimizes connectivity scores, $( C_i)$, assigned to species to maximize the likelihood of recreating the observed network pattern. The likelihood \( L \) for this model is:

$$
L = \prod_{i,j} \left( C_i \cdot C_j \right)^{A_{ij}} \left( 1 - C_i \cdot C_j \right)^{1 - A_{ij}}
$$

where $( C_i)$ is the connectivity score of species $(i)$.


- **3. Matching-Trait Model:**

The matching-trait model assumes that species interactions are not random but based on trait differences. It optimizes parameters by scoring interactions along one or multiple latent-trait axes. The interaction probability $( P_{ij})$ between species \( i \) and \( j \) is determined by:

$$
P_{ij} = f(T_i, T_j)
$$

where $( T_i)$ and $( T_j)$ are trait vectors for species \( i \) and \( j \), and \( f \) is a function measuring trait similarity or compatibility.



- **4. Matching-Centrality Model:**

The matching-centrality model combines the approaches of the connectance and matching-trait models. It optimizes both species connectivity scores and latent trait axes. The likelihood \( L \) for this model is:

$$
L = \prod_{i,j} \left( C_i \cdot C_j \cdot f(T_i, T_j) \right)^{A_{ij}} \left( 1 - C_i \cdot C_j \cdot f(T_i, T_j) \right)^{1 - A_{ij}}
$$

Following the guidelines established by Poisot (2023), Youden's J statistic was used to compare model performance. Youden's J is a metric of model informedness that balances sensitivity and specificity, calculated as:

$$
J = \text{Sensitivity} + \text{Specificity} - 1
$$




# Data Accessibility Statement

Data is open-source, digitally available at their respective sources. 

# Conflict of Interest Statement

The authors declare no conflict of interest.